A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-01-21 paper.Rmd
done B-P part

The BP procedure consists of three steps:

@@

```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950."}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, 
     col = "red", lwd = 2, xlab = "Age (days)", lty = 2)
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red", lty = 2)
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

@@

```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed postneonatal mortality data."}
ginfdat <- age.window(infdat, c(28, 365))
ginfdat$enter <- g(ginfdat$enter)
ginfdat$exit <- g(ginfdat$exit)
grs <- risksets(with(ginfdat, Surv(enter, exit, event)), members = FALSE)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = ginfdat)
ser <- plot(fit, fn = "sur", fig = FALSE)
##plot(fit, col = "blue", axes = FALSE)
cum <- fit$hazards[[1]]
x <- cum[, 1]
ch <- cumsum(cum[, 2])
plot(x, ch, type = "s", axes = FALSE, col = "blue", xlab = "Age (days)", ylab = "")
axis(1, at = c(g(28), g(183), g(365)), labels = c(28, 183, 365))
axis(2, las = 1)
abline(h = 0)
box()
x <- ser[[1]][[1]][, 1]
y <- ser[[1]][[1]][, 2]
lines(x, 1 - y, col = "red", lty = 2)
legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeouis-Pichat)"), col = c("blue", "red"),
       lty = c(1, 2))
```

@@

The way to disentangle the endogenous and exogenous componets is to start with a plot 
of the cumulative hazard function (Figure \@ref(fig:separ)).

@@

axis(1, at = c(0, g(28), g(365)), labels = c(0, 28, 365), pos = 0)
axis(2, las = 1, pos = 0, at = c(0, 0.02, 0.04, 0.06), labels = c(0, 2, 4, 6))
#box()
abline(v = 0)
texhaz <- g(365) * fitex$hazards[1, 1]
lines(c(0, g(365)), c(0, texhaz), col = "red", lwd = 1.5)
atte <- c(0.0000, 0.0763, texhaz)
axis(4, at = atte, labels = 100 * round(atte, 4), las = 1, cex = 0.5,
     pos = g(365))
abline(v = g(365))
text(100, 0.03, "Exogenous", col = "red", srt = 22, cex = 0.8)
text(100, 0.053, "Endogenous plus exogenous", col = "blue", srt = 22, cex = 0.8)
axis(3, at =  c(0, g(28), g(365)), labels = c(0, 28, 365))
```

So we can see from Figure \@ref(fig:separ) that of the total mortality, 
`r round(5.27 / 7.63 * 100, 1)` per cent is exogenous and the rest is endogenous.
AS always, this requires that we trust the Bourgeois-Pichat model.

@@

Remains to be written. Feels like two papers, one about Bourgeouis-Pichat and one
about climate and weather. Maybe discuss exogeneity/endogeneity?

>>>>

The B-P procedure consists of three steps:

@@

```{r ggplot2, fig.cap = "B-B plots with and without time transform, Umeå-Skellefteå 1895-1950.", fig.height = 4}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, 
     col = "red", lwd = 2, xlab = "Age (days)", lty = 2)
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red", lty = 2)
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

@@

```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed postneonatal mortality data."}
ginfdat <- age.window(infdat, c(28, 365))
ginfdat$enter <- g(ginfdat$enter)
ginfdat$exit <- g(ginfdat$exit)
grs <- risksets(with(ginfdat, Surv(enter, exit, event)), members = FALSE)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = ginfdat)
ser <- plot(fit, fn = "sur", fig = FALSE)
##plot(fit, col = "blue", axes = FALSE)
cum <- fit$hazards[[1]]
x <- cum[, 1]
ch <- cumsum(cum[, 2])
plot(x, ch, type = "s", axes = FALSE, col = "blue", xlab = "Age (days)", ylab = "")
axis(1, at = c(g(28), g(183), g(365)), labels = c(28, 183, 365))
axis(2, las = 1)
abline(h = 0)
box()
x <- ser[[1]][[1]][, 1]
y <- ser[[1]][[1]][, 2]
lines(x, 1 - y, col = "red", lty = 2)
legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeois-Pichat)"), col = c("blue", "red"),
       lty = c(1, 2), cex = 0.7)
```

@@

The way to disentangle the endogenous and exogenous components is to start with a plot 
of the *cumulative* hazard function and add the exogenous 
part from the postneonatal estimate, see Figure \@ref(fig:separ).

@@

axis(1, at = c(0, g(28), g(365)), labels = c(0, 28, 365), pos = 0)
axis(2, las = 1, pos = 0, at = c(0, 0.0096, 0.0338), labels = c(0, 0.96, 3.38))
#box()
abline(v = 0)
texhaz <- g(365) * fitex$hazards[1, 1]
lines(c(0, g(365)), c(0, texhaz), col = "red", lwd = 1.5)
atte <- c(0.0000, 0.0763, texhaz)
axis(4, at = atte, labels = 100 * round(atte, 4), las = 1, cex = 0.5,
     pos = g(365))
abline(v = g(365))
text(100, 0.03, "Exogenous", col = "red", srt = 22, cex = 0.8)
text(100, 0.053, "Endogenous plus exogenous", col = "blue", srt = 22, cex = 0.8)
axis(3, at =  c(0, g(28), g(365)), labels = c(0, 28, 365))
lines(c(0, g(28)), c(0.0096, 0.0096), lty = 3)
lines(c(0, g(28)), c(0.0338, 0.0338), lty = 3)
abline(v = g(28))
```

So we can see from Figure \@ref(fig:separ) (vertical axis on the right hand side) that of the total mortality, 
`r round(5.27 / 7.63 * 100)` per cent is exogenous and the rest is endogenous.
Similarly, of the neonatal mortality, `r round(0.96 / 3.38 * 100)` per cent is
exogenous and the rest is endogenous (vertical axis on the left hand side).
As always, this requires that we trust the Bourgeois-Pichat model.

@@

Remains to be written. Feels like two papers, one about Bourgeois-Pichat and one
about climate and weather. Maybe discuss exogeneity/endogeneity?
